---
title: "Candidate Gene Analysis, Regression method"
author: "Jared J. Homola"
date: "7/31/2020"
output: html_document
---

These analyses examine the relationship between expression levels and viral titer for a set of genes that we hypothesize have a relationship with viral infection. The overall strategy is to:  
  
  1. Identify a list of candidate genes based on the literature and previous analyses.    
  2. Subset our overall set of transcripts to include only those with an annotation containing the name of the candidate gene in the SwissProt gene name, SwissProt gene function, KEGG pathway, or EggNOG gene orthology annotations.  
  3. Manually filter search results to remove non-target inclusions  
  4. Remove transcripts with no variation in expression for each comparison (e.g., bursa at transcript-level)  
  5. Perform linear regression for each candidate gene/transcript separate using a model of expression ~ virus shedding group + sex + weight on day 55 + age with the sequencing pool as a random effect. Regressions are performed on the entire set, as well as just on groups I1, I3, I5, and I14.
  6. Correct the p-values for multiple testing using a false discovery rate approach with adjusted p-values (i.e., q-value) cutoff of 0.05.  
  7. Plot the results  
  
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("G:/Shared drives/RNA Seq Supershedder Project/BWTE DE manuscript/")
load("candidateGeneAnalysis_regression.Rws")
```

```{r message = FALSE, warning = FALSE, echo = FALSE, results = 'hide'}
## Load packages and data
library(tidyverse)
library(kableExtra)
library(edgeR)
library(ggrepel)

## Load data
annot <- read.delim("G:/Shared drives/RNA Seq Supershedder Project/BWTE DE manuscript/extData/Trinotate.csv", header = TRUE, sep = "\t")
cnt.trans <- read.table("G:/Shared drives/RNA Seq Supershedder Project/BWTE DE manuscript/extData/rsem.isoform.counts.matrix", header = TRUE)
cnt.gene <- read.table("G:/Shared drives/RNA Seq Supershedder Project/BWTE DE manuscript/extData/rsem.gene.counts.matrix", header = TRUE)
covars <- read.csv("G:/Shared drives/RNA Seq Supershedder Project/BWTE DE manuscript/extData/BWTE54_SSgroup_Raw_Pool.csv", header = TRUE)
targets <- read_delim("G:/Shared drives/RNA Seq Supershedder Project/BWTE DE manuscript/extData/aPrioriTranscripts_V2.csv", delim = ",") %>%
  filter(include == "Yes" | include == "yes")

## Clean data
cnt.bursa.gene <- cnt.gene %>%
  select(-alignEstimateAbundance_BWTE_Ileum_36_S50,
         -alignEstimateAbundance_BWTE_Bursa_36_S31,-alignEstimateAbundance_BWTE_Ileum_19_S35,
         -alignEstimateAbundance_BWTE_Bursa_19_S14) %>%
  rownames_to_column("gene") %>%
  separate(gene, into = c(NA, "pt1", "pt2", "gene", NA)) %>%
  unite(gene, pt1, pt2, gene) %>%
  column_to_rownames("gene") %>%
  select(contains("Bursa"))

cnt.ileum.gene <- cnt.gene %>%
  select(-alignEstimateAbundance_BWTE_Ileum_36_S50,
         -alignEstimateAbundance_BWTE_Bursa_36_S31,-alignEstimateAbundance_BWTE_Ileum_19_S35,
         -alignEstimateAbundance_BWTE_Bursa_19_S14) %>%
  rownames_to_column("gene") %>%
  separate(gene, into = c(NA, "pt1", "pt2", "gene", NA)) %>%
  unite(gene, pt1, pt2, gene) %>%
  column_to_rownames("gene") %>%
  select(contains("Ileum"))

cnt.bursa.trans <- cnt.trans %>%
  select(-alignEstimateAbundance_BWTE_Ileum_36_S50,
         -alignEstimateAbundance_BWTE_Bursa_36_S31,-alignEstimateAbundance_BWTE_Ileum_19_S35,
         -alignEstimateAbundance_BWTE_Bursa_19_S14) %>%
  rownames_to_column("transcript") %>%
  separate(transcript, into = c(NA, "pt1", "pt2", "gene", "isoform")) %>%
  unite(transcript, pt1, pt2, gene, isoform) %>%
  column_to_rownames("transcript") %>%
  select(contains("Bursa"))

cnt.ileum.trans <- cnt.trans %>%
  select(-alignEstimateAbundance_BWTE_Ileum_36_S50,
         -alignEstimateAbundance_BWTE_Bursa_36_S31,-alignEstimateAbundance_BWTE_Ileum_19_S35,
         -alignEstimateAbundance_BWTE_Bursa_19_S14) %>%
  rownames_to_column("transcript") %>%
  separate(transcript, into = c(NA, "pt1", "pt2", "gene", "isoform")) %>%
  unite(transcript, pt1, pt2, gene, isoform) %>%
  column_to_rownames("transcript") %>%
  select(contains("Ileum"))

covars <- covars %>%
  filter(!bird %in% c("36", "19")) %>%
  arrange(bird) %>%
  mutate(group = str_remove(group, "-"))

annot <- annot %>%
  separate(transcript_id, into = c(NA, "pt1", "pt2", "gene", "isoform")) %>%
  unite(gene_id, pt1, pt2, gene, remove = FALSE) %>%
  unite(transcript_id, pt1, pt2, gene, isoform) %>% 
  filter(transcript_id %in% targets$transcript_id | gene_id %in% targets$gene_id)

#### Calculate log(CPM) and assemble master DFs ####
#Convert to DGEList object
dge.bursa.trans <- DGEList(counts=cnt.bursa.trans)
dge.bursa.gene <- DGEList(counts=cnt.bursa.gene)
dge.ileum.trans <- DGEList(counts=cnt.ileum.trans)
dge.ileum.gene <- DGEList(counts=cnt.ileum.gene)

#CPM and log-CPM
lcpm.bursa.trans <- cpm(dge.bursa.trans, log = TRUE)
lcpm.bursa.gene <- cpm(dge.bursa.gene, log = TRUE)
lcpm.ileum.trans <- cpm(dge.ileum.trans, log = TRUE)
lcpm.ileum.gene <- cpm(dge.ileum.gene, log = TRUE)

## Master lcpm tibs
# Trans
lcpm.bursa.tmp <- lcpm.bursa.trans %>%
  as_tibble(rownames = NA) %>%
  rownames_to_column("transcript")
lcpm.ileum.tmp <- lcpm.ileum.trans %>%
  as_tibble(rownames = NA) %>%
  rownames_to_column("transcript")

lcpm.trans <- lcpm.bursa.tmp %>%
  full_join(lcpm.ileum.tmp) %>%
  replace(., is.na(.), "0") %>%
  filter(transcript %in% targets$transcript_id) %>%
  pivot_longer(cols = contains("_"),
               names_to = "sample",
               values_to = "lcpm") %>%
  separate(sample, into = c(NA, NA, "tissue", "bird", NA)) %>%
  mutate(bird = as.integer(bird)) %>%
  left_join(covars, by = "bird") %>%
  mutate(levelGT = "transcript", identifier = transcript) %>%
  select(identifier, levelGT, tissue, bird, lcpm, virus.sac, group, virus.dpi1, virus.dpi2, virus.dpi3, virus.dpi4, virus.dpi5) %>%
  mutate(log.virus.sac = log10(virus.sac+1)) %>%
  pivot_longer(cols = virus.dpi1:virus.dpi5,
               names_to = "dpi",
               values_to = "titer") %>%
  group_by(identifier, bird, levelGT, tissue, lcpm, log.virus.sac, group) %>%
  summarize(log.meanTiter1to5 = log10(mean(titer, na.rm = TRUE)))

# Gene
lcpm.bursa.tmp <- lcpm.bursa.gene %>%
  as_tibble(rownames = NA) %>%
  rownames_to_column("gene")
lcpm.ileum.tmp <- lcpm.ileum.gene %>%
  as_tibble(rownames = NA) %>%
  rownames_to_column("gene")

lcpm.all <- lcpm.bursa.tmp %>%
  full_join(lcpm.ileum.tmp) %>%
  replace(., is.na(.), "0")  %>%
  filter(gene %in% targets$gene_id) %>%
  pivot_longer(cols = contains("_"),
               names_to = "sample",
               values_to = "lcpm") %>%
  separate(sample, into = c(NA, NA, "tissue", "bird", NA)) %>%
  mutate(bird = as.integer(bird)) %>%
  left_join(covars, by = "bird") %>%
  mutate(levelGT = "gene", identifier = gene) %>%
  select(identifier, levelGT, tissue, bird, lcpm, virus.sac, group, virus.dpi1, virus.dpi2, virus.dpi3, virus.dpi4, virus.dpi5) %>%
  mutate(log.virus.sac = log10(virus.sac+1)) %>%
  pivot_longer(cols = virus.dpi1:virus.dpi5,
               names_to = "dpi",
               values_to = "titer") %>%
  group_by(identifier, bird, levelGT, tissue, lcpm, log.virus.sac, group) %>%
  summarize(log.meanTiter1to5 = log10(mean(titer + 1, na.rm = TRUE))) %>%
  bind_rows(lcpm.trans) %>% 
  mutate(group = recode(group,
                        C1 = "Ctl",
                        C14 = "Ctl")) %>% 
  mutate(group = factor(group, 
                          levels = c("Ctl", "I1", "I3", "I5", "I14")))

#### Gene query database ####
annot.all <- annot %>%
  select(transcript_id,
         gene_id,
         sprot_Top_BLASTX_hit,
         sprot_Top_BLASTP_hit,
         gene_ontology_BLASTX,
         gene_ontology_BLASTP,
         Kegg,
         eggnog,
         Pfam) %>%
  separate(sprot_Top_BLASTX_hit, into = c("sprot_geneName_BlastX", NA, NA, NA, NA, "sprot2", NA), "\\^") %>%
  separate(sprot2, sep = "=", into = c(NA, "sprot_geneFunction_BlastX")) %>%
  separate(sprot_geneFunction_BlastX, sep = ";", into = c("sprot_geneFunction_BlastX", NA)) %>%
  separate(sprot_Top_BLASTP_hit, into = c("sprot_geneName_BlastP", NA, NA, NA, NA, "sprot2", NA), "\\^") %>%
  separate(sprot2, sep = "=", into = c(NA, "sprot_geneFunction_BlastP")) %>%
  separate(sprot_geneFunction_BlastP, sep = ";", into = c("sprot_geneFunction_BlastP", NA)) %>%
  separate(gene_ontology_BLASTX, sep = "\\`", into = paste("GO_BlastX", 1:5, sep = "_"), extra = "drop", fill = "right") %>%
  separate(gene_ontology_BLASTP, sep = "\\`", into = paste("GO_BlastP", 1:5, sep = "_"), extra = "drop", fill = "right") %>%
  separate(Pfam, sep = "\\`", into = paste("Pfam", 1:5, sep = "_"), extra = "drop", fill = "right") %>%
  select(transcript_id, gene_id, sprot_geneName_BlastX, sprot_geneFunction_BlastX) %>% 
  as_tibble()
```

## Summarize results
```{r, message = FALSE, warning = FALSE, echo = FALSE}
finalMatrix.BG.sig <- finalMatrix.BG %>%
  filter(group == "all", result.type == "pval") %>%
  mutate(adj.p.value = p.adjust(results.value, method='fdr', n = nrow(.))) %>%
  mutate(comparison = "BG")

finalMatrix.BT.sig <- finalMatrix.BT %>%
  filter(group == "all", result.type == "pval") %>%
  mutate(adj.p.value = p.adjust(results.value, method='fdr', n = nrow(.))) %>%
  mutate(comparison = "BT")

finalMatrix.IG.sig <- finalMatrix.IG %>%
  filter(group == "all", result.type == "pval") %>%
  mutate(adj.p.value = p.adjust(results.value, method='fdr', n = nrow(.))) %>%
  mutate(comparison = "IG")

finalMatrix.IT.sig <- finalMatrix.IT %>%
  filter(group == "all", result.type == "pval") %>%
  mutate(adj.p.value = p.adjust(results.value, method='fdr', n = nrow(.))) %>%
  mutate(comparison = "IT") 

finalMatrix.all.sig <- bind_rows(finalMatrix.BG.sig,
                                 finalMatrix.BT.sig,
                                 finalMatrix.IG.sig,
                                 finalMatrix.IT.sig) %>% 
  filter(adj.p.value < 0.05) %>% 
  select(-results.value, -result.type) %>% 
  mutate(result.type = "adj.p.value",
         results.value = adj.p.value) %>% 
  select(result.type, identifier, results.value, comparison) %>% 
  mutate(tempID = paste0(identifier, "_", comparison))

finalMatrix.BG <- finalMatrix.BG %>% mutate(comparison = "BG")
finalMatrix.BT <- finalMatrix.BT %>% mutate(comparison = "BT")
finalMatrix.IG <- finalMatrix.IG %>% mutate(comparison = "IG")
finalMatrix.IT <- finalMatrix.IT %>% mutate(comparison = "IT")

allResults.Sig <- bind_rows(finalMatrix.BG, 
                            finalMatrix.BT,
                            finalMatrix.IG,
                            finalMatrix.IT) %>% 
  filter(group == "all") %>% 
  select(-group) %>% 
  mutate(tempID = paste0(identifier, "_", comparison)) %>% 
  filter(tempID %in% finalMatrix.all.sig$tempID) %>% 
  bind_rows(finalMatrix.all.sig) %>% 
  select(-tempID) %>% 
  filter(result.type != "pval") %>% 
  print(n = 400)
  
results.table <- allResults.Sig %>% 
  pivot_wider(names_from = result.type, values_from = results.value) %>% 
  select(identifier, comparison, adj.p.value, adj.r.squared, estimate)

kable(results.table) %>%
  kable_styling("striped", full_width = F)
```

```{r, message = FALSE, warning = FALSE, echo = FALSE}
results.all <- bind_rows(finalMatrix.BG,
                         finalMatrix.BT,
                         finalMatrix.IG,
                         finalMatrix.IT)

```


## Functions for analyzing and plotting data
```{r, message = FALSE, warning = FALSE, echo = FALSE}
correlationPlot.mean <- function(target, targetTissue, targetLevel, ...) {
  annotation <- annot.all %>%
    filter(transcript_id == target | gene_id == target) %>%
    filter(sprot_geneName_BlastX != ".") %>%
    select(sprot_geneName_BlastX) %>%
    slice(1)
  
  legendData <- results.all %>% 
    filter(identifier == target) %>% 
    mutate(tissue = if_else(comparison == "BG" | comparison == "BT", "Bursa", "Ileum")) %>% 
    mutate(level = if_else(comparison == "BG" | comparison == "IG", "gene", "transcript")) %>% 
    filter(tissue == targetTissue, level == targetLevel)
  
  plot <- lcpm.all %>%
    filter(levelGT == targetLevel,
           tissue == targetTissue) %>% 
    filter(identifier == target) %>% 
    ggplot(aes(x = log.virus.sac, y = lcpm)) +
    geom_point(aes(shape = group, color = group), size = 5) +
    geom_smooth(method = "lm", color = "black") +
    ylab("Log2(Counts per million)") +
    xlab("Log10(Sacrifice day viral titer + 1)") +
    scale_shape_manual(values=c(13, 15:18),
                       name = "Group", 
                       labels = c("Ctl",
                                  paste0("I1: ","P-value = ", round(filter(legendData, group == "I1")$results.value, 3)), 
                                  paste0("I3: ","P-value = ", round(filter(legendData, group == "I3")$results.value, 3)), 
                                  paste0("I5: ","P-value = ", round(filter(legendData, group == "I5")$results.value, 3)),
                                  paste0("I14: ","P-value = ", round(filter(legendData, group == "I14")$results.value, 3)))) +
    scale_color_manual(values = c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442"),
                       name = "Group", 
                       labels = c("Ctl",
                                  paste0("I1: ","P-value = ", round(filter(legendData, group == "I1")$results.value, 3)), 
                                  paste0("I3: ","P-value = ", round(filter(legendData, group == "I3")$results.value, 3)), 
                                  paste0("I5: ","P-value = ", round(filter(legendData, group == "I5")$results.value, 3)),
                                  paste0("I14: ","P-value = ", round(filter(legendData, group == "I14")$results.value, 3)))) +
    guides(shape = guide_legend(override.aes = list(size = 5))) +
    theme_classic() +
    labs(title=paste0(annotation[1], " - ", target),
         subtitle=paste0(targetTissue, " - ", targetLevel,
                         "; Adj. p = ", round(targetInfo$adj.p.value, 5),
                         "; Pearson correlation coefficent = ", round(targetInfo$results.value, 3)))
  print(plot)
}

```

## Plotting significant results for sacrifice day titer
```{r, message = FALSE, warning = FALSE, echo = FALSE}
for(z in (1:nrow(results.table))) {
  targetInfo <- results.table %>% 
    slice(z) %>% 
    separate(comparison, c("metric", "tissueLevel")) %>% 
    mutate(tissue = if_else(tissueLevel == "BG" | tissueLevel == "BT", "Bursa", "Ileum")) %>% 
    mutate(level = if_else(tissueLevel == "BG" | tissueLevel == "IG", "gene", "transcript"))

  if(targetInfo$metric == "sac") {
    correlationPlot.sac(targetInfo$target, targetInfo$tissue, targetInfo$level)
  }
  else {
    correlationPlot.mean(targetInfo$target, targetInfo$tissue, targetInfo$level)
  }
}
```
