---
title: "Late stage infection - Bursa - Gene"
author: "Jared J. Homola"
date: "6/16/2020"
output: html_document
---
This is an analysis of differential expression at the gene level between shedding rate groupings based on average shedding rates for infected blue-winged teal bursa samples.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
## Load packages and data
library(limma)
library(edgeR)
library(Glimma)
library(gplots)
library(ggrepel)
library(RColorBrewer)
library(kableExtra)
library(gridExtra)
library(topGO)
library(tidyverse)

## Load data
annot <- read.delim("G:/Shared drives/RNA Seq Supershedder Project/BWTE DE manuscript/extData/Trinotate.csv", header = TRUE, sep = "\t")
cnt <- read.table("G:/Shared drives/RNA Seq Supershedder Project/BWTE DE manuscript/extData/rsem.gene.counts.matrix", header = TRUE)
covars <- read.csv("G:/Shared drives/RNA Seq Supershedder Project/BWTE DE manuscript/extData/BWTE54_SSgroup_Raw_Pool.csv", header = TRUE)
```


### Clean data
Bird 36 is missing a measurement for the weight covariate and bird 19 had improper tissue-based grouping, suggesting a potential sample mix up. They are both removed from the analysis. We also rename the genes, dropping the "Trinity_" prefix.
```{r, warning = FALSE, message = FALSE}
cnt <- cnt %>%
  select(
    -alignEstimateAbundance_BWTE_Ileum_36_S50,
    -alignEstimateAbundance_BWTE_Bursa_36_S31,-alignEstimateAbundance_BWTE_Ileum_19_S35,
    -alignEstimateAbundance_BWTE_Bursa_19_S14) %>%
  rownames_to_column("gene") %>%
  separate(gene, into = c(NA, "pt1", "pt2", "gene")) %>%
  unite(gene, pt1, pt2, gene) %>%
  column_to_rownames("gene") %>% 
  select(contains("Bursa"))

covars <- covars %>%
  filter(!bird %in% c("36", "19")) %>%
  arrange(bird) %>%
  mutate(group = str_remove(group, "-"))

annot <- annot %>%
  separate(transcript_id, into = c(NA, "pt1", "pt2", "gene", "isoform")) %>%
  unite(transcript_id, pt1, pt2, gene, isoform) %>% 
  separate(gene_id, into = c(NA, "pt1", "pt2", "gene")) %>%
  unite(gene_id, pt1, pt2, gene)
```


### Set up model effects
```{r, warning = FALSE, message = FALSE}
bird <- as.factor(covars$bird)
sex <- as.factor(covars$sex)
age <- as.numeric(covars$age)
weight <- covars$wt_55
group <- recode(covars$group, C1 = "Ctl", C14 = "Ctl") %>% 
  as.factor()
pool <- as.factor(covars$Pool.Bursa)
ssgroup.virus.avg <- as.factor(covars$SSgroup.virus.avg)

covars.tib <- data.frame(bird,
                    sex,
                    age,
                    weight,
                    group,
                    pool,
                    ssgroup.virus.avg) %>% 
  as_tibble()
```

### Prep data for analysis
Here we standardize expression levels across samples of varying depth by using the counts per million (CPM) and trimmed mean of M-values (TMM) methods. We also remove any genes that are expressed at few than 0.5 CPM in at least 25% of individuals. 
```{r, warning = FALSE, message = FALSE}
#Convert to DGEList object
dge <- DGEList(counts=cnt)
dge$genes <- annot[match(rownames(dge$counts), annot$gene_id),]

#CPM and log-CPM
cpm <- cpm(dge)
lcpm <- cpm(dge, log = TRUE)

#### Retain only genes expressed at >0.5 CPM in at least 25% of individuals
table(rowSums(dge$counts==0) == length(cnt[1,]))
keep.exprs <- rowSums(cpm>0.5) >= length(cnt[1,])/4
sum(keep.exprs)

dge <- dge[keep.exprs,, keep.lib.sizes=FALSE]
table(rowSums(dge$counts==0) == length(cnt[1,]))
dim(dge)

#TMM normalization
dge <- calcNormFactors(dge, method = "TMM")
```

### Subset data
Following normalization, we subset the data based on infection groups.
```{r, warning = FALSE, message = FALSE}
#Assign groups in DGElist
dge$samples$group <- covars.tib$group

### Divide out I5
dge.I5 <- dge[,grep('I5', dge$samples$group)]
covars.I5 <- covars.tib %>% filter(group == "I5")
```


# Differential expression analysis
```{r, warning = FALSE, message = FALSE,}
## Establish design matrix: I5
design.I5 = model.matrix(~ 0 +
                        factor(covars.I5$ssgroup.virus.avg) +
                        covars.I5$age +
                        covars.I5$sex + 
                        covars.I5$weight + 
                        covars.I5$pool)

colnames(design.I5) <- c("HIGH", "LOW", "MODERATE",
                         "age", "sexM", "weight", "pool2", "pool3")


contr.matrix.I5 = makeContrasts(
  LvM = LOW - MODERATE,
  MvH = MODERATE - HIGH,
  LvH = LOW - HIGH,
  levels = colnames(design.I5))

## Mean-variance trend plots
v.I5 <- voomWithQualityWeights(dge.I5, design.I5, plot = FALSE)

#Fitting models
vfit.I5 <- lmFit(v.I5, design.I5) 
vfit.I5 <- contrasts.fit(vfit.I5, contrasts = contr.matrix.I5)
tfit.I5 <- treat(vfit.I5, lfc = 1.0)

## Identify DE genes
dt.I5 <- decideTests(tfit.I5, p.value = 0.1, adjust.method = "fdr")
dt.tib.I5 <- as_tibble(dt.I5, rownames = NA) %>% 
  rownames_to_column("gene") %>% 
  mutate_at(vars(starts_with("L")), as.numeric) %>% 
  mutate_at(vars(starts_with("M")), as.numeric) %>% 
  rename(LvM.I5 = LvM,
         MvH.I5 = MvH,
         LvH.I5 = LvH)

dt.tib <- dt.tib.I5 %>% 
  mutate_at(vars(starts_with("L")), as.numeric) %>% 
  mutate_at(vars(starts_with("M")), as.numeric) %>% 
  filter_at(vars(LvM.I5:LvH.I5), any_vars(. != 0))
```

### Numbers of DE genes
```{r}
allResults <- as.data.frame(summary(dt.I5)) %>% 
  filter(Var1 != "NotSig")

kable(allResults) %>%
  kable_styling("striped", full_width = F)
```

### Volcano plot
```{r, warning = FALSE, message = FALSE, fig.height = 10, fig.width = 10, fig.align = "center"}
tmp1 <- topTreat(tfit.I5, coef = 1, n = Inf)
results1 <- mutate(tmp1, sig=ifelse(tmp1$adj.P.Val<0.1, "Sig", "Not Sig"))
p1 <- ggplot(results1, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Low vs. Moderate shedding") +
  ylab("-Log10(Adjusted p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")


tmp2 <- topTreat(tfit.I5, coef = 2, n = Inf)
results2 <- mutate(tmp2, sig=ifelse(tmp2$adj.P.Val<0.1, "Sig", "Not Sig"))
p2 <- ggplot(results2, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Moderate vs. High shedding") +
  ylab("-Log10(Adjusted p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp3 <- topTreat(tfit.I5, coef = 3, n = Inf)
results3 <- mutate(tmp3, sig=ifelse(tmp3$adj.P.Val<0.1, "Sig", "Not Sig"))
p3 <- ggplot(results3, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Low vs. High shedding") +
  ylab("-Log10(Adjusted p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

grid.arrange(p1, p2, p3, nrow = 1)
```


### Annotations for differentially expressed genes
```{r message=FALSE, warning=FALSE}
annot %>%
  select(
    gene_id,
    sprot_Top_BLASTX_hit
  ) %>%
  rename(gene = gene_id) %>% 
  inner_join(dt.tib) %>% 
  separate(sprot_Top_BLASTX_hit, into = c("sprot1", NA, NA, NA, NA, "sprot2", NA), "\\^") %>% 
  separate(sprot2, sep = "=", into = c(NA, "SwissProt_GeneFunction")) %>% 
  separate(sprot1, sep = "_", into = c("SwissProt_GeneName", NA)) %>% 
  mutate_at(.vars = vars(LvM.I5:LvH.I5),
            .funs = ~ ifelse(. == 1, "up", .)) %>% 
  mutate_at(.vars = vars(LvM.I5:LvH.I5),
            .funs = ~ ifelse(. == -1, "down", .)) %>% 
  filter_at(vars(LvM.I5:LvH.I5), any_vars(. != 0)) %>% 
  distinct(gene, .keep_all = TRUE) %>% 
  select(-SwissProt_GeneFunction) %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

#### Gene functions
```{r message=FALSE, warning=FALSE}
annot %>%
  select(
    gene_id,
    sprot_Top_BLASTX_hit,
    Kegg) %>%
  rename(gene = gene_id) %>% 
  inner_join(dt.tib) %>% 
  separate(sprot_Top_BLASTX_hit, into = c("sprot1", NA, NA, NA, NA, "sprot2", NA), "\\^") %>% 
  separate(sprot2, sep = "=", into = c(NA, "SwissProt_GeneFunction")) %>% 
  separate(sprot1, sep = "_", into = c("SwissProt_GeneName", NA)) %>% 
  separate(Kegg, into = c("KEGG_ID", "KO_ID"), sep = "\\`") %>% 
  distinct(gene, SwissProt_GeneName, KEGG_ID, .keep_all = TRUE) %>% 
  select(gene, SwissProt_GeneName, SwissProt_GeneFunction, KEGG_ID, KO_ID) %>% 
  filter(SwissProt_GeneName != ".") %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

### Gene Ontology Enrichment Analysis
```{r message=FALSE, warning=FALSE}
topGO.mappings <- readMappings("G:/Shared drives/RNA Seq Supershedder Project/BWTE DE manuscript/extData/GOdbGene.txt", sep = "\t", IDsep = ",")

goEnrich <- function(targetComp, ...) {
  DE.results <- dt.tib %>% filter(!!sym(targetComp) != 0)
  
  all.genes <- sort(unique(as.character(rownames(dge$counts))))
  int.genes <- DE.results$gene
  int.genes <- factor(as.integer(all.genes %in% int.genes))
  names(int.genes) = all.genes
  
  go.obj.BP <- new("topGOdata", ontology='BP',
              allGenes = int.genes,
              annot = annFUN.gene2GO,
              nodeSize = 5,
              gene2GO = topGO.mappings)

  go.obj.MF <- new("topGOdata", ontology='MF',
                 allGenes = int.genes,
                 annot = annFUN.gene2GO,
                 nodeSize = 5,
                 gene2GO = topGO.mappings)

  go.obj.CC <- new("topGOdata", ontology='CC',
                 allGenes = int.genes,
                 annot = annFUN.gene2GO,
                 nodeSize = 5,
                 gene2GO = topGO.mappings)
  
  results.BP <- runTest(go.obj.BP, algorithm = "elim", statistic = "fisher")

  results.tab.BP <- GenTable(object = go.obj.BP, 
                          elimFisher = results.BP, 
                          topNodes = 50) %>% 
    as_tibble() %>% 
    mutate(pVal = as.numeric(elimFisher)) %>% 
    filter(pVal < 0.01) %>% 
    mutate(Domain = "BP") %>% 
    mutate(Comparison = targetComp)
  
  results.MF <- runTest(go.obj.MF, algorithm = "elim", statistic = "fisher")
  
  results.tab.MF <- GenTable(object = go.obj.MF, 
                          elimFisher = results.MF, 
                          topNodes = 50) %>% 
    as_tibble() %>% 
    mutate(pVal = as.numeric(elimFisher)) %>% 
    filter(pVal < 0.01) %>% 
    mutate(Domain = "MF") %>% 
    mutate(Comparison = targetComp)
  
  results.CC <- runTest(go.obj.CC, algorithm = "elim", statistic = "fisher")
  
  results.tab.CC <- GenTable(object = go.obj.CC, 
                          elimFisher = results.CC, 
                          topNodes = 50) %>% 
    as_tibble() %>% 
    mutate(pVal = as.numeric(elimFisher)) %>% 
    filter(pVal < 0.01) %>% 
    mutate(Domain = "CC") %>% 
    mutate(Comparison = targetComp)
  
  rbind(results.tab.BP,
    results.tab.MF,
    results.tab.CC)
}

results <- list()

for(z in unique(names(dt.tib)[-1])) {
  if (dt.tib %>% 
      filter(!!sym(z) != 0) %>% 
      nrow(.) > 0)
  results[[length(results)+1]] <- goEnrich(z)
}

results.tib <- bind_rows(results, .id = "column_label") %>%
  dplyr::select(-column_label) %>% 
  arrange(GO.ID, Comparison)

kable(results.tib) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```
