---
title: "Tissue type - Transcript"
author: "Jared J. Homola"
date: "4/12/2020"
output: html_document
---
This is an analysis of differential expression at the transcript level between bursa and ileum tissues from blue-winged teals. The goal of this analysis is to evaluate whether the tissues should be analyzed separately.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning = FALSE, message = FALSE}
## Load packages and data
library(limma)
library(edgeR)
library(Glimma)
library(gplots)
library(RColorBrewer)
library(kableExtra)
library(tidyverse)

setwd("G:/Shared drives/RNA Seq Supershedder Project/BWTE DE manuscript/bwteDE")

## Load data
annot <- read.delim("./extData/Trinotate.csv", header = TRUE, sep = "\t")
cnt <- read.table("./extData/rsem.isoform.counts.matrix", header = TRUE)
covars <- read.csv("./extData/BWTE54_SSgroup_Raw_Pool.csv", header = TRUE)
```


### Clean data
Bird 36 is missing a measurement for the weight covariate and bird 19 had improper tissue-based grouping, suggesting a potential sample mix up. They are both removed from the analysis. We also rename the transcripts, dropping the "Trinity_" prefix.
```{r, warning = FALSE, message = FALSE}
cnt <- cnt %>%
  select(
    -alignEstimateAbundance_BWTE_Ileum_36_S50,
    -alignEstimateAbundance_BWTE_Bursa_36_S31,-alignEstimateAbundance_BWTE_Ileum_19_S35,
    -alignEstimateAbundance_BWTE_Bursa_19_S14) %>%
  rownames_to_column("transcript") %>%
  separate(transcript, into = c(NA, "pt1", "pt2", "gene", "isoform")) %>%
  unite(transcript, pt1, pt2, gene, isoform) %>%
  column_to_rownames("transcript")

covars <- covars %>%
  filter(!bird %in% c("36", "19")) %>%
  arrange(bird) %>%
  mutate(group = str_remove(group, "-"))

annot <- annot %>%
  separate(transcript_id, into = c(NA, "pt1", "pt2", "gene", "isoform")) %>%
  unite(transcript_id, pt1, pt2, gene, isoform)
```


### Set up model effects
```{r, warning = FALSE, message = FALSE}
tissue <- as.factor(c(rep("Bursa", ncol(cnt)/2), rep("Ileum", ncol(cnt)/2)))
bird <- as.factor(rep(covars$bird, 2))
sex <- as.factor(rep(covars$sex, 2))
age <- as.numeric(rep(covars$age, 2))
weight <- rep(covars$wt_55, 2)
group <- as.factor(rep(covars$group, 2))
pool <- as.factor(c(covars$Pool.Bursa, covars$Pool.Ileum))
```

### Prep data for analysis
Here we standardize expression levels across samples of varying depth by using the counts per million (CPM) and trimmed mean of M-values (TMM) methods. We also remove any transcripts that are expressed at few than 0.5 CPM in at least 25% of individuals.
```{r, warning = FALSE, message = FALSE}
#Convert to DGEList object
dge <- DGEList(counts=cnt)
dge$genes <- annot[match(rownames(dge$counts), annot$transcript_id),]

#CPM and log-CPM
cpm <- cpm(dge)
lcpm <- cpm(dge, log = TRUE)

#### Retain only transcripts expressed at >0.5 CPM in at least 25% of individuals
table(rowSums(dge$counts==0) == length(cnt[1,]))
keep.exprs <- rowSums(cpm>0.5) >= length(cnt[1,])/4
sum(keep.exprs)

dge <- dge[keep.exprs,, keep.lib.sizes=FALSE]
table(rowSums(dge$counts==0) == length(cnt[1,]))
dim(dge)

#TMM normalization
dge <- calcNormFactors(dge, method = "TMM")
```

### Plot MDS
```{r}
## Sex and tissue
col.group = as.numeric(as.factor(tissue))
plotMDS(lcpm, labels = paste0(bird,"_",tissue), col = col.group, cex = 0.75)
title(main="Tissue type")
```

# Differential expression analysis
### Establish design & contrast matrices
```{r}
## Establish design matrix
design = model.matrix(~ 0 +
                        tissue*group +
                        age +
                        sex + 
                        weight + 
                        pool) 

colnames(design) <- gsub("tissue", "", colnames(design))
colnames(design) <- gsub(":", "_", colnames(design))


contr.matrix = makeContrasts(
  tissue = Bursa - Ileum,
  levels = colnames(design))
```

### Mean-variance trend and sample weight plots
```{r}
v <- voomWithQualityWeights(dge, design, plot = TRUE)
```

### Fitting the model
```{r}
vfit <- lmFit(v, design) 
vfit <- contrasts.fit(vfit, contrasts = contr.matrix)
tfit <- treat(vfit, lfc = 1)
plotSA(tfit)
```

### Count DE transcripts
For a transcript to be considered differentially expressed, we require a p-value of 0.05 with a false discovery rate correction and a log fold count difference of 1.
```{r}
dt <- decideTests(tfit, p.value = 0.1, adjust.method = "fdr")
summary(dt) %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F)
```

### Histogram of P-values
```{r}
hist(tfit$p.value, main = "P-values comparing groups (eBayes)")
```
